Efficient evaluation of threshold queries of derived fields in a numerical simulation database

### 2014-08-28 paper.tex
Some changes to the figures.

%Why is it hard? (E.g., why do naive approaches fail?)
The evaluation of threshold queries of derived fields requires a kernel computation at each point on the grid in order to compute the field to be thresholded.
The value of the derived field at each location depends on all the values at locations within the kernel of computation. 
In order to evaluate such a query for a given time-step all of the data for that time-step must be read and each data point 
will appear in $N$ kernels of computation, where $N$ is the size of the kernel. The evaluation is thus both data and computation intensive.
It is impractical to materialize all possible derived fields due to the large size of the datasets and the limits of available storage. Therefore, threshold 
queries have to be evaluated on-demand from the stored data.
Database, operating and file system caches are effective at speeding up access to the large amounts of data stored on disk. However, as was
shown by Lopez et al. \cite{Lopez} this might not be sufficient for some interactive applications. Moreover, even if the data are available in the cache the
computation associated with the derived field still needs to be performed for each point on the grid as results of previous computations are not cached.

@@

%Why hasn't it been solved before? (Or, what's wrong with previous proposed solutions? How does mine differ?)
Data-intensive architectures and compute clusters built from commodity hardware rely on parallel I/O to multiple disks and high network bandwidth
to achieve high throughput. Such systems have only recently been deployed for the storage of large numerical simulation 
datasets. The virtual laboratories built on these systems often do not support
all of the functionality that scientists are interested in. Database systems support rollup queries, including top-$k$ queries, but in most cases these
queries apply only simple linear score functions on the attribute values of individual records. 
We present a method for the efficient evaluation of threshold queries over fields derived from the raw vector or scalar fields of the numerical simulation 
stored in the database. Our method computes and operates on derived fields, which are derived by performing kernel computations over the array data 
stored in the database.
Our method makes effective use of the cluster resources and achieves high throughput and scalability.

%What are the key components of my approach and results? Also include any specific limitations.
We exploit the parallelism 
available in the cluster by means of data-parallel execution of the computations. Each computation is distributed across the nodes of the cluster and
is executed in parallel by several threads within a database node by means of creating multiple connections to each node. Moreover we introduce an 
application-aware cache for query results. We build on the idea of an application-aware cache introduced in \cite{Lopez}, but rather than just data 
objects we cache the actual results of the computation, which leads to query performance improvement of up to X times. In that regard the cache
that we use shares more similarities with the idea of semantic caching \cite{Ren}.

- The introduction of an application-aware cache improves the running time of queries substantially if there is a cache hit and there is 
little downside to storing query results in the cache.

- The evaluation scales with the amount of parallelism available in the cluster.

@@

The central part of the evaluation strategy for threshold queries that we have developed is the introduction of an application-aware cache
for query results. The results of these queries are small compared to the amount of data that need to be examined and the results can be used to
answer subsequent queries as long as they are within the same region and specify the same or higher threshold. Caching the query results
preserves the computational effort in addition to substantially reducing I/O. We do not have to derive the requested field from the raw data for 
queries that hit the cache. This results in a substantial improvement in query performance as we only have to scan
a small set of data and do not need to perform any additional computation.
\kk{Cite results for top-$k$ queries in databases and distributed systems. Remark that non of these strategies operate on large array datasets 
and they do not operate on derived fields, where each value of the derived field depends on multiple values from the original field and requires a kernel
computation accessing multiple data points in a kernel around the target location.}

@@

The central part of the strategy that we have developed for the evaluation of threshold queries of derived fields is the application-aware cache, which
stores the results of these queries. We first evaluate the overhead associated with the introduction and maintenance of the cache. Fig. \ref{fig:cache_overhead}
compares the execution time of queries in the absence of a cache with the execution time of the same queries, which interrogate the cache first.
For these experiments we requested
the locations with norms of the vorticity above thresholds at different levels. For the first set the threshold was set high and only $\sim$4,300 points 
(or 0.0004\% of all points) were above the threshold. For the second set a medium threshold was chosen and $\sim$87,000 points 
(or 0.0081\% of all points) were above the threshold. Finally, a low threshold was chosen for the last set with $\sim$900,000 points 
(or 0.0847\% of all points) above the threshold. For each set a random
time-step was chosen and the queries were run against that time-step. The measurements were taken from the point of view of the end user. 
The execution times were averaged over 10 runs and cache entries for the particular time-step were dropped before each run. 
We utilized 4 threads per database
node for the evaluation of each query. As we can see from the results the overhead associated with querying the cache first is small, less than 5\% and within
the margin of error. Additionally, the method shows stable running time across different time-steps and threshold levels
\kk{Consider including a figure for a representative time-step showing the PDF of the vorticity field
to show how the vorticity values are distributed and to get an understanding for the different threshold levels.}

@@

We observer a two times speedup when going from a single thread per node to two threads per node. The speedup diminishes to 1.3 times when going 
to 4 threads and little speedup is 
observed with 8 threads per node. This is due to several factors. Each node has 2 CPUs and a total of 8 cores, which need to run operating system and other SQL Server
processes as well. The experiments were run on the live production database nodes, which were also servicing other user queries. This is also exhibited in
the large variance that is seen in the running time of the queries. Additionally, SQL
Server already utilizes parallelism to perform most of the I/O. Finally, due to the data-parallel approach to the evaluation of these queries the data atoms along
the boundary of each region will be requested by multiple threads. Nevertheless, running with 4 threads per node is nearly 2.5 times faster when compared to
running a single thread.
\kk{Should we have scale-out experiments as well?}

@@

Zhao et al. propose an algorithm for the processing of top-$k$ queries in large-scale distributed environments called BRANCA \cite{Zhao}. They build on the 
idea of semantic caching \cite{Ren} and make use of brach caches, which store results of previous top-$k$ queries with respect to the data stored on each server.
The caching mecahnism that we use is similar in that regard, but the queries that are evaluated in our system operate on derived fields, which are computed
at each location by accessing data from a surrounding region. The queries described in \cite{Zhao} operate over the attributes of individual records only using
simple linear score functions.

>>>>

%Why is it hard? (E.g., why do naive approaches fail?)
The evaluation of threshold queries of derived fields requires a kernel computation at each point on the grid in order to compute the field to be thresholded.
The value of the derived field at each location depends on all the values at locations within the kernel of computation. 
In order to evaluate such a query for a given time-step all of the data for that time-step must be read and each data point 
will appear in $N$ kernels of computation, where $N$ is the size of the kernel. The evaluation is thus both data and computation intensive.
It is impractical to materialize all possible derived fields due to the large size of the datasets and the limits of available storage. Therefore, the derived fields 
and threshold queries on them have to be evaluated on-demand from the stored data.
Database, operating and file system caches are effective at speeding up access to the large amounts of data stored on disk. However, as was
shown by Lopez et al. \cite{Lopez} this might not be sufficient for some interactive applications. Moreover, even if the data are available in the cache the
computation associated with the derived field still needs to be performed for each point on the grid as results of previous computations are not cached.

@@

%Why hasn't it been solved before? (Or, what's wrong with previous proposed solutions? How does mine differ?)
Data-intensive architectures and compute clusters built from commodity hardware rely on parallel I/O to multiple disks and high network bandwidth
to achieve high throughput. Such systems have only recently been deployed for the storage of large numerical simulation 
datasets. The virtual laboratories built on these systems often do not support
all of the functionality that scientists are interested in. Database systems support rollup queries, including top-$k$ queries, but in most cases these
queries apply only simple linear score functions on the attribute values of individual records. 
We present a method for the efficient evaluation of threshold queries over fields derived from the raw vector or scalar fields of the numerical simulation 
stored in the database. The derived fields, which our method computes and operates on are derived by performing kernel computations over the array data 
stored in the database.
Our method makes effective use of the cluster resources and achieves high throughput and scalability.

%What are the key components of my approach and results? Also include any specific limitations.
We exploit the parallelism 
available in the cluster by means of data-parallel execution of the computations. Each computation is distributed across the nodes of the cluster and
is executed in parallel by several threads within a database node by means of creating multiple connections to each node. Moreover we introduce an 
application-aware cache for query results. We build on the idea of an application-aware cache introduced by Lopez et al. \cite{Lopez} and more broadly 
on the concept of semantic caching \cite{Ren}. Rather than caching just data as is the case in system caches and the tree-cache described by Lopez et al.
we cache the actual results of the computation, which leads to query performance improvement of up to X times.

@@

The central part of the evaluation strategy for threshold queries that we have developed is the introduction of an application-aware cache
for query results. The results of these queries are small compared to the amount of data that need to be examined and the results can be used to
answer subsequent queries as long as they are within the same region and specify the same or higher threshold. Caching the query results
preserves the computational effort in addition to substantially reducing I/O. We do not have to derive the requested field from the raw data for 
queries that hit the cache. This results in a substantial improvement in query performance as we only have to scan
a small set of data and do not need to perform any additional computation.
%\kk{Cite results for top-$k$ queries in databases and distributed systems. Remark that non of these strategies operate on large array datasets 
%and they do not operate on derived fields, where each value of the derived field depends on multiple values from the original field and requires a kernel
%computation accessing multiple data points in a kernel around the target location.}

@@

The central part of the strategy that we have developed for the evaluation of threshold queries of derived fields is the application-aware cache, which
stores the results of these queries. We first evaluate the overhead associated with the introduction and maintenance of the cache. Fig. \ref{fig:cache_overhead}
compares the execution time of queries in the absence of a cache with the execution time of the same queries, which interrogate the cache first.
For these experiments we requested
the locations with norms of the vorticity above thresholds at different levels. For the first set the threshold was set high and only $\sim$4,300 points 
(or 0.0004\% of all points) were above the threshold. For the second set a medium threshold was chosen and $\sim$87,000 points 
(or 0.0081\% of all points) were above the threshold. Finally, a low threshold was chosen for the last set with $\sim$900,000 points 
(or 0.0847\% of all points) above the threshold. For each set a random
time-step was chosen and the queries were run against that time-step. The measurements were taken from the point of view of the end user. 
The execution times were averaged over 10 runs and cache entries for the particular time-step were dropped before each run. 
We utilized 4 threads per database
node for the evaluation of each query. As we can see from the results the overhead associated with querying the cache first is small, less than 3\% and within
the margin of error. Additionally, the method shows stable running time across different time-steps and threshold levels
\kk{Consider including a figure for a representative time-step showing the PDF of the vorticity field
to show how the vorticity values are distributed and to get an understanding for the different threshold levels.}

@@

We observer nearly a two times speedup when going from a single thread per node to two threads per node. The speedup diminishes to 1.4 times when going 
to 4 threads and little speedup is 
observed with 8 threads per node. This is due to several factors. Each node has 2 CPUs and a total of 8 cores, which need to run operating system and other SQL Server
processes as well. The experiments were run on the live production database nodes, which were also servicing other user queries. This is also exhibited in
the large variance that is seen in the running time of the queries. Additionally, SQL
Server already utilizes parallelism to perform most of the I/O. Finally, due to the data-parallel approach to the evaluation of these queries the data atoms along
the boundary of each region will be requested by multiple threads. Nevertheless, running with 4 threads per node is nearly 2.6 times faster when compared to
running a single thread.
\kk{Should we have scale-out experiments as well?}

@@

Zhao et al. propose an algorithm for the processing of top-$k$ queries in large-scale distributed environments called BRANCA \cite{Zhao}. They build on the 
idea of semantic caching \cite{Ren} and make use of branch caches, which store results of previous top-$k$ queries with respect to the data stored on each server.
The caching mecahnism that we use is similar in that regard, but the queries that are evaluated in our system operate on derived fields, which are computed
at each location by accessing data from a surrounding region. The queries described in \cite{Zhao} operate over the attributes of individual records only using
simple linear score functions.
